Short and Mechanized Logical Relation for Dependent Type Theory

### 2024-02-21 paper/paper.tex
some work on the introduction. not done yet

\section{Introduction}
In this paper, we present a short and mechanized consistency proof by
logical relation for \lang{}, a Martin-Löf style dependent type theory
with large elimination, a full predicative universe hierarchy, an
intensional identity type and a boolean base type, both of which
equipped with dependent elimination forms. % The mechanization is
% extremely short and involves less than 1000 lines of manually written
% Coq code.

Our goal is to show the fundamental technique required to carry out proof by logical
relation in a dependently typed setting where logical relation can no
longer be defined as a simple recursive function over the structure of
types. Compared to the many mechanized proofs by logical relation for
predicative dependent type
theory~\citep{nbeincoq,decagda,martin-lof-a-la-coq,anand2014towards},
whose proof developments range from 20,000 to 400,000 lines of code,
our Coq development is the most lightweight at less than 1,000 lines.

The big difference in the size of the developments does not
necessarily imply that our proof technique leads to a more concise
proof due to the differing expressiveness of the languages and the
different metatheoretic results being established. For
example, \citet{anand2014towards} mechanizes the metatheory of
Nuprl~\citep{constable1986implementing}, which, on top of all the
features that \lang{} supports, includes W-types and partial types and
requires a PER semantics to model its extensional typed
equality. The object language from \citet{decagda} does not support
identity types and only has one predicative universe, but additionally
supports $\Sigma$ types. However, for a researcher who wishes to learn
the underlying techniques for mechanized logical relation for
dependent type theory, a small development like ours is much easier to
navigate and understand. Furthermore, we show how our simple
consistency proof can be easily extended to show the existence of
$\beta\eta$-normal form for well-typed open \emph{and} closed terms,
giving us a metatheoretic result almost as strong as the one from
\citet{decagda} at the moderate cost of around 500 lines of extra code.

We present our definitions and proofs in set theory notation and
terminology rather than in Coq syntax to make the material accessible
to readers with general mathematical background.
However, notation aside, the typeset representation is
purposefully made to be closer to our mechanization so less massaging
of the definitions is required to encode our proofs in a proof
assistant.
The close correspondence between a typeset proof and a mechanized
proof comes at the potential risk that the artifacts specific to
theorem provers may become visible in the presented proof.
However, for proofs by logical relation for
dependent types, we find our approach advantageous for a few reasons.
First, unlike properties that are derivable through
syntactic means, proof by logical relation is demanding on the
strength of the metatheory. An informal proof that attempts to be
agnostic or ambiguous about the underlying metatheory % may struggle to even
% convey the validity of their definition and
requires substantial
effort from the reader to even understand whether it is theoretically
possible to encode such proofs in a proof assistant.
Second, while a proof assistant may reject perfectly valid
definitions, it helps us keep our definition precise and
unambiguous. A precise definition is crucial as the reasoning
principles we are allowed to use is directly tied to the definition,
which is sometimes obscured by concise notations that diverge too far
from a formal proof.
Finally, our approach does not result in a verbose proof on paper, thanks to
the already short mechanization. On top of that, We make conservative
simplifications to our typeset proof such as avoiding the clunky
syntax for well-founded recursion in Coq, but we keep the overall
structure of our typeset proofs consistent with our mechanization.
The consistency enables us to label at the footnote each lemma
directly to their counterpart in the proof script for the readers to
reference and validate.

@@

The complexity of applying proof by logical relation to dependent types stem
from the fact that the logical relation is much harder to define. In
simply typed languages, the logical relation is defined as a recursive
function over the type $[[A]]$. In dependent types, the type
$[[A]]$ can take the form $[[(\ var 0) Bool]]$. To assign meaning to
this type, we need to first reduce it to $[[Bool]]$. However, we
cannot write a function that performs the reduction because we do not
know the termination of well-typed terms a priori. As a result, we
define the logical relation as an inductively defined relation,
reminiscent of how we specify the reduction graph of a partial
function; the functionality of the relation can later be recovered in Lemma~\ref{lemma:logreldeter}.

\label{sec:logreldep}
\begin{figure}[h]
\drules[I]{$[[Interp I i A S]]$}{Logical Relation}{Void, Bool, Eq, Pi, Set, Red}
\caption{Logical relation for \lang}
\label{fig:logrel}
\end{figure}
The logical relation for \lang{}, which takes the form $[[Interp I i A
S]]$, is defined as an inductively generated relation (Figure~\ref{fig:logrel}).
% The logical relation takes the form $[[Interp I i A S]]$.
Metavariables $[[A]]$ and $[[i]]$ stand for terms and natural
numbers respectively, as introduced earlier in
Figure~\ref{fig:syntax}.
\scw{Many introductory texts define the relation as a recursive function
over type structure, or step-indices. You use an inductive relation instead, why?}
\scw{Can we view this inductive relation as the graph of the partial function
that is defined recursively over types?}
\scw{What is this form extensible too? impredicative quantification? recursive
types? }
\scw{Is it worth observing here that this definition is not over sets of typed
terms. That it characterizes all terms that look like booleans (i.e. evaluate to
true or false) or all terms that look like proofs (i.e. evaluate to refl). The
fact that there is no connection between p and a and b in the I-Eq case is strange
looking. Need to explain.  }
The metavariables $[[I]]$ and $[[S]]$ are
sets with the following signatures.
\begin{equation*}
  \begin{split}
    [[I]] &\in [[ { j | j < i  } ->  PowerSet STm ]] \\
    [[S]] &\in [[PowerSet STm]]
  \end{split}
\end{equation*}
The notation $[[PowerSet STm]]$ denotes the powerset of the set of
\lang{} terms.
The function $[[I]]$ is a family of sets of terms indexed by
natural numbers strictly less than the parameter $[[i]]$, which
represents the current universe level.  In \rref{I-Set}, function
$[[I]]$ is used to define the meaning of
universes that are strictly smaller than the current level $[[i]]$. The
restriction $[[j < i]]$ in \rref{I-Set} is crucial for our system to
be predicative and we will explain the reason shortly.

@@

To derive consistency, it suffices to restrict $[[S]]$ to the set of
terms that reduce to closed terms. \Rref{I-Void, I-Bool} are
unsurprising; when considering only closed terms, the empty type
should not be inhabited and therefore corresponds to the empty set,
whereas the boolean type is semantically inhabited by terms that
evaluate to the boolean values $[[true]]$ or $[[false]]$. \Rref{I-Eq}
says that an equality type $[[a ~ b : A]]$ corresonds to the
set of terms that evaluate to $[[refl]]$ when $[[a <=> b]]$ holds and
otherwise corresponds to the empty set. Side conditions like $[[a <=>
b]]$ are typically required for indexed types, of which equality types
are an instance. \Rref{I-Par} enables us to reduce our types in order
to assign meanings. Recall the type expression $[[(\ var 0)
Bool]]$. \Rref{I-Par} says to know that
$[[Interp I i  (\ var 0) Bool S ]]$ for some $[[S]]$, it suffices to
show that $[[Interp I i Bool S]]$ since $[[(\var 0) Bool =>
Bool]]$. The derivation that $[[Interp I i (\ var 0) Bool { a | a =>+
  true \/ a =>+ false }]]$ therefore follows by composing \rref{I-Par}
and \rref{I-Bool}.

\Rref{I-Pi} is the most interesting. To explain what it means, we
need to perform a few steps of transformations to its precondition.
The precondition consists of a
mysterious relation $[[R]]$ over the set $[[S]]$, an interpretation
of the type $[[A]]$, and returns a subset of terms. $[[R]]$ is further
subject to the following two constraints:
\begin{itemize}
\item $[[forall a, (# exists S0  , ( a , S0 ) in R #)]]$
\item $[[forall a, forall S0, (# ( a , S0 ) in R implies Interp I i B { a } S0 #)]]$
\end{itemize}
The first condition allows us to induce a function $[[F in S ->
PowerSet STm]]$ such that $[[(a , F a) in R]]$. Then we can easily
show that $[[forall a, (# a in S implies Interp I i B { a } F a #)]]$.

For the other direction, given a function $[[F]]$ satisfying the
property $[[forall a, (# a in S implies Interp I i B { a } F a #) ]]$, it
is possible to go in the other direction and define a relation $[[R]]=[[
{ ( a , F a )  | a in S }]]$. It is easy to verify that $[[R]]$ satisfy
the the constraint required in the precondition of \rref{I-Pi}.

Furthermore, the skolemization process allows us to conclude the
following equivalence:
\[ \exists F, [[forall a, (# a in S implies Interp I i B { a } F a #)]] \iff [[forall a, (# a in S implies (# exists S0 , Interp I i B {a} S0 #) #)]]  \]
% Ignoring the content below the horizontal bar of
% the derivation, we claim the statements from the precondition are equivalent to
% the conjunction of the following statements:
% \begin{itemize}
% \item $[[Interp I i A S]]$
% \item $\forall [[a]], \text{ if }[[a in S]]\text{, then } \exists
%   [[S0]]\text{, } [[Interp I i B { a } S0 ]] $
% \end{itemize}
% In general, let $R$ be a binary relation over the sets $[[A]]$ and
% $[[B]]$, it easy to verify the following equivalence, which is used to
% justify the skolemization process~\citep{skolemization}.
% \scw{Is this true constructively? If not, is that ok?}
% \[\forall a \in A, \exists b \in B \text{ such that } (a,b) \in R
%   \iff \exists F \in A \rightarrow B \text, \forall a \in A, (a, F(a))
%   \in R\]
% The left-hand side is equivalent to the right-hand side, but has no
% mentioning of any functions.
The precondition of \rref{I-Pi} therefore can be formulated as the
following statements:
\begin{itemize}
\item $[[Interp I i A S]]$
\item $[[forall a, (# a in S implies (# exists S0 , Interp I i B {a} S0 #) #)]]$
\end{itemize}
From this alternative formulation, we can more easily tell what is
happening in \rref{I-Pi}: the
function type $[[Pi A B]]$ has an interpretation if its input
type $[[A]]$ can be
interpreted as some set $[[S]]$, and for all terms $[[a in S]]$, the
type $[[B {a}]]$, obtained by substituing $[[a]]$ into the output type
$[[B]]$, has some semantic interpretation.

If we look at $[[Interp I i Pi A B { b | forall a, forall S0, (# ( a , S0 ) in R implies b a in S0 #) }]]$, the conclusion of \rref{I-Pi}, we see why the
presentation with an explicit $[[R in S * PowerSet STm]]$ is
useful. We want to state the fact that $[[B {a}]]$ has a semantic
interpretation of some set $[[S0]]$, but we also want to refer to the
set $[[S0]]$ when we talk about the set that $[[Pi A B]]$ corresponds
to. With the relation $[[R]]$ in scope, we can refer to the
witness sets $[[S0]]$ such that $[[Interp I i B {a} S0]]$ for $[[a in S]]$.
With the alternative representation that is free of the relation $[[R]]$, we
might want to reformulate our logical relation in the following more
concise but equivalent form, whose admissibility follows directly from
\rref{I-Pi}.
\begin{center}
  \drule[]{I-PiAlt}
\end{center}
\begin{lemma}
  \label{lemma:piintroalt}
  \Rref{I-PiAlt} is admissible.
\end{lemma}
\begin{proof}
Immediate from \rref{I-Pi} with $[[R]]$ instantiated to the relation $[[{ (a
, S0 ) | a in S , Interp I i B { a } S0 }]]$.
\end{proof}

Unfortunately, \rref{I-PiAlt} not only violates the syntactic strict
positivity constraint required in proof assistants, but
is genuinely non-monotone when written as an endofunction over the
domain of relations.
Intuitively, the failure of monotonicity stems from the fact
that the witness picked in the precondition is not necessarily the
same witness being referred to in the post condition. While it might
be possible to restrict the domain with additional constraints such as
functionality and inversion properties, we opt for our current
formulation of \rref{I-Pi} so we immediately obtain a
well-defined inductive relation and a usable induction principle. The
slight disadvantage of \rref{I-Pi} is that we need to construct the
function $[[R]]$ each time we apply it, though this is mitigated by
the immediate admissibility of \rref{I-PiAlt}.

@@

First, we prove a family of simple properties, which we refer to as
inversion principles for our logical relation. Given $[[Interp I i A
S]]$ where $[[A]]$ is in some head form such as $[[Bool]]$ or $[[Pi A0
B0]]$, the inversion lemma allows us to say something about the set
$[[S]]$. Its proof is simple, but we sketch out the case for
functions to help readers confirm their understanding of \rref{I-Pi}.
\begin{lemma}[Inversion of the logical relation]
  \label{lemma:interpinv}\leavevmode
  \begin{enumerate}
  \item\footnote{\dotv{semtyping.v}{InterpExt\_Void\_inv}} If $[[Interp I i Void S]]$, then $[[S = emptyset]]$.
  \item\footnote{\dotv{semtyping.v}{InterpExt\_Bool\_inv}} If $[[Interp I i Bool S]]$, then $[[S = { a | a =>+ true \/ a =>+ false   }]]$.
  \item\footnote{\dotv{semtyping.v}{InterpExt\_Eq\_inv}} If $[[Interp I i a ~ b : A S]]$, then $[[S = { p | p =>+ refl , a <=> b  }]]$.
  \item\footnote{\dotv{semtyping.v}{InterpExt\_Fun\_inv}} If $[[Interp I i Pi A B S1]]$, then there exists $[[S]],[[R]]$ such that:
    \begin{itemize}
    \item $[[Interp I i A S ]]$
    \item $[[R in S * PowerSet STm]]$
    \item $[[forall a, (# a in S implies (# exists S0 , (a , S0) in R #) #)]]$
    \item $[[forall a, forall S0, (# ( a , S0 ) in R implies Interp I i B { a } S0 #)]]$
    \item $[[S1 = { b | forall a, forall S0, (# ( a , S0 ) in R implies b a in S0 #) }]]$
    \end{itemize}
  \item\footnote{\dotv{semtyping.v}{InterpExt\_Univ\_inv}} If $[[Interp I i Set j S]]$, then $[[j < i]]$ and $[[S = I j]]$.
  \end{enumerate}
\end{lemma}
\begin{proof}
  As mentioned earlier, we only show the inversion property for the
  function type.
  We start by inducting over the derivation of $[[Interp I i Pi A B S]]$. There
  are only two possible cases we need to consider.
  \begin{description}
  \item[\Rref{I-Pi}:] Immediate.
  \item[\Rref{I-Red}:] We are given that $[[Interp I i Pi A B S1]]$.
    We know that there exists some $[[A0]]$ and
    $[[B0]]$ such that $[[Pi A B => Pi A0 B0]]$ and $[[Interp I i Pi
    A0 B0 S1]]$. From the
    induction hypothesis, there exists $[[S]]$ and $[[R]]$ such that :
    \begin{itemize}
    \item $[[Interp I i A0 S ]]$
    \item $[[R in S * PowerSet STm]]$
    \item $[[forall a, (# a in S implies (# exists S0 , (a , S0) in R #) #)]]$
    \item $[[forall a, forall S0, (# ( a , S0 ) in R implies Interp I i B0 { a } S0 #)]]$
    \item $[[S1 = { b | forall a, forall S0, (# ( a , S0 ) in R implies b a in S0 #) }]]$
    \end{itemize}
    By inverting the derivation of $[[Pi A B => Pi A0 B0]]$, we derive $[[A => A0]]$ and
    $[[B => B0]]$. By Lemma~\ref{lemma:parsubst}, we have $[[B {a} => B0 {a} ]]$ for all
    $[[a]]$. As a result, by \rref{I-Red}, the same $[[S]]$ and
    $[[R]]$ additionally satisfies the following properties:
    \begin{itemize}
    \item $[[Interp I i A S ]]$
    \item $[[forall a, forall S0, (# ( a , S0 ) in R implies Interp I i B { a } S0 #)]]$
    \end{itemize}
    These properties are exactly what we need to finish the proof.
  \end{description}
\end{proof}

\Rref{I-Red} bakes into the logical relation the backward preservation
property. That is, given $[[Interp I i A S]]$, if $[[B =>+ A]]$, then
$[[Interp I i B S]]$ also holds. The following property shows that
preservation holds in the usual forward direction too.
\begin{lemma}[Preservation of the logical relation\footnote{\dotv{semtyping.v}{InterpExt\_preservation}}]
  \label{lemma:interppreservation}
  If $[[Interp I i A S]]$ and $[[A => B]]$, then $[[Interp I i B S]]$.
\end{lemma}
\begin{proof}
  We carry out the proof by induction over the derivation of $[[Interp
  I i A S]]$.

  The only interesting case is \Rref{I-Red}. Given that
  $[[A => B0]]$ and $[[Interp I i B0 S]]$, we need to show
  for all  $[[B]]$ such that $[[A => B1]]$, we have $[[Interp I i B1
  S]]$. By the diamond property of parallel reduction
  (Lemma~\ref{lemma:pardiamond}), there exists some term $[[B]]$ such
  that $[[B0 => B]]$ and $[[B1 => B]]$. By the induction hypothesis,
  we deduce $[[Interp I i B S]]$ from $[[B0 => B]]$ and $[[Interp I i
  B0 S]]$. By \rref{I-Par} and $[[B1 => B]]$, we conclude that
  $[[Interp I i B S]]$.

  The remaining cases all fall from induction hypotheses and basic
  properties about convertibility and parallel reduction we have
  established in Section~\ref{sec:spec}.
\end{proof}
From Lemma~\ref{lemma:interppreservation} and \rref{I-Red}, we can easily
derive the following corollary that two convertible types can always interpret
into the same set. We adopt the terminology from \citet{martin-lof-a-la-coq}
and refer to this property as irrelevance.
\begin{corollary}[Irrelevance of logical relation\footnote{\dotv{semtyping.v}{InterpUnivN\_Coherent}}]
  \label{lemma:logrelcoherence}
  If $[[Interp I i A S]]$ and $[[A <=> B]]$, then $[[Interp I i B S]]$.
\end{corollary}
% \scw{Need a transition here}

Since the definition of our logical relation is an inductive relation,
it is not immediately obvious why each type $[[A]]$ can only uniquely
corresponds to one set $[[S]]$. The following lemma shows that our
logical relation is indeed functional.
\begin{lemma}[Logical relation is functional\footnote{\dotv{semtyping.v}{InterpExt\_deterministic}}]
  \label{lemma:logreldeter}
  If $[[Interp I i A S0]]$ and $[[Interp I i A S1]]$, then $[[S0 = S1]]$.
\end{lemma}
\begin{proof}
  The proof proceeds by induction over the derivation of the first
  premise $[[Interp I i A S0]]$.
  All cases that are not \rref{I-Red} follow immediately from the
  Lemma~\ref{lemma:interpinv}.

>>>>

\section{Introduction}
In this paper, we present a \emph{short}, \emph{mechanized} proof of logical
consistency for \lang{}, a Martin-Löf style dependent type theory with large
eliminations, a full predicative universe hierarchy, an intensional identity
type and a boolean base type, both of which are equipped with dependent
elimination forms.

Our goal with this work is to demonstrate how the proof technique of
\emph{logical relations} can be applied in dependently-typed setting.
Tutorial material on this topic~\cite{lr-tutorials} is primarily focussed on
languages with simple or polymorphic types. In that context, logical relations
can be define as simple recursive functions over the structure of types, or
(in the case of recursive types) defined over the evaluation steps of the
computation.

Neither of these techniques make sense for predicative dependent type theory,
so a young researcher might be excused for thinking that logical
relations-based arguments are not available.  \scw{Well, actually, we did use
  a step-indexed relation for Trellys\ldots} But this is not the case. Recent
authors have developed tour-de-force mechanizations for the metatheory of
modern proof
assistants~\citep{nbeincoq,decagda,martin-lof-a-la-coq,anand2014towards}, and
have used logical-relations-based reasoning as part of their
arguments. However, at the same time, these definitions have been overshadowed
by their magnitude of the surrounding developments. Because these proofs show
diverse results about real systems and algorithms, these developments range in
size from 20,000 to 400,000 lines of code, and the techniques are not
accessible to a casual reader.

Our proof, designed to capture the essence of this proof technique in this
setting, requires less than 1000 lines of code. We have achieved this
significant reduction through a number of means: the careful selection of the
features that we include in the object type type system and the results that
we prove about it, in addition to the judicious use of automation.  Our
language is small, but includes enough to be illustrative. For example, we
eschew inductive datatypes or W-types, the but we do include propositionality
equality and booleans in order to capture the challenges presented by indexed
types and dependent pattern matching. We do not show the decidability of type
checking, nor do we develop a PER semantics, but we do demonstrate how our
consistency proof can be extended (at a moderate cost of 500 lines of code) to
show the existence of $\beta\eta$-normal form for well-typed open \emph{and}
closed terms. We include a full predicative universe hiercharchy and
type-level computation, to demonstrate the logical strength of the
approach. \scw{Say that we don't have impredicativity, but neither does anyone
  else?}
%%
%% SCW: save specific comparisons with other systems for later
%%
% The big difference in the size of the developments does not
% necessarily imply that our proof technique leads to a more concise
% proof due to the differing expressiveness of the languages and the
% different metatheoretic results being established. For
% example, \citet{anand2014towards} mechanizes the metatheory of
% Nuprl~\citep{constable1986implementing}, which, on top of all the
% features that \lang{} supports, includes W-types and partial types and
% requires a PER semantics to model its extensional typed
% equality. The object language from \citet{decagda} does not support
% identity types and only has one predicative universe, but additionally
% supports $\Sigma$ types. However, for a researcher who wishes to learn
% the underlying techniques for mechanized logical relation for
% dependent type theory, a small development like ours is much easier to
% navigate and understand. Furthermore, we show how our simple
% consistency proof can be easily extended to show the existence of
% $\beta\eta$-normal form for well-typed open \emph{and} closed terms,
% giving us a metatheoretic result almost as strong as the one from
% \citet{decagda} at the moderate cost of around 500 lines of extra code.

The result of our work is a development that an interested researcher can
navigate and understand. We accompany this mechanized proof with an informal
description, presented here using set theory notation and terminology so that
the material is accessible to readers with general mathematical background,
That said, our explanations do not stray too far away from our proof scripts.
We label each lemma directly to their counterpart in the proof script,
anticipating that readers may wish to reference and validate our formalism.
Our typeset representation are purposefully made to directly follow our
mechanical proofs, while avoiding, as much as possible, the introduction 
of artifacts that are specific to certain theorem prover.

Not only does this connection aid readers that wish to, like us, adopt proof
assistants for their day-to-day use. We also find that this precision is
important for conveying the proof techique itself. Unlike properties that are
derivable through syntactic means, proofs by logical relation make demands on
the strength of the logic in which they are expressed. An informal proof that
attempts to be agnostic or ambiguous about the underlying
metatheory % may struggle to even
% convey the validity of their definition and
requires substantial
effort from the reader to even understand whether it is theoretically
possible to encode such proofs in a given ambient logic, such as those found 
in modern proof assistants.

% The close correspondence between a typeset proof and a mechanized
% proof comes at the potential risk that the artifacts specific to
% theorem provers may become visible in the presented proof.
%However, for proofs by logical relation for
%dependent types, we find our approach advantageous for a few reasons.
Furthermore, while a proof assistant may reject perfectly valid definitions,
it helps us keep our definition precise and unambiguous. A precise definition
is crucial as the reasoning principles we are allowed to use is directly tied
to the definition, which is sometimes obscured by concise notations that
diverge too far from a formal proof.

Finally, our approach does not result in a verbose proof on paper, thanks to
the already short mechanization. While we do make conservative simplifications
to our typeset proof, such as avoiding the clunky syntax for well-founded
recursion in Coq, we keep the overall structure of our typeset proofs
consistent with our mechanization.

@@

\label{sec:logreldep}
\begin{figure}[h]
\drules[I]{$[[Interp I i A S]]$}{Logical Relation}{Void, Bool, Eq, Pi, Set, Red}
\caption{Logical relation for \lang}
\label{fig:logrel}
\end{figure}
The logical relation for \lang{}, which takes the form $[[Interp I i A
S]]$, is defined as an inductively generated relation (Figure~\ref{fig:logrel}).
% The logical relation takes the form $[[Interp I i A S]]$.
Metavariables $[[A]]$ and $[[i]]$ stand for terms and natural
numbers respectively, as introduced earlier in
Figure~\ref{fig:syntax}.
\scw{Many introductory texts define the relation as a recursive function
over type structure, or step-indices. You use an inductive relation instead, why?}
\scw{Can we view this inductive relation as the graph of the partial function
that is defined recursively over types?}
\scw{What is this form extensible too? impredicative quantification? recursive
types? }
\scw{Is it worth observing here that this definition is not over sets of typed
terms. That it characterizes all terms that look like booleans (i.e. evaluate to
true or false) or all terms that look like proofs (i.e. evaluate to refl). The
fact that there is no connection between p and a and b in the I-Eq case is strange
looking. Need to explain.  }

The metavariables $[[I]]$ and $[[S]]$ are
sets with the following signatures.
\begin{equation*}
  \begin{split}
    [[I]] &\in [[ { j | j < i  } ->  PowerSet STm ]] \\
    [[S]] &\in [[PowerSet STm]]
  \end{split}
\end{equation*}
The notation $[[PowerSet STm]]$ denotes the powerset of the set of
\lang{} terms.
The function $[[I]]$ is a family of sets of terms indexed by
natural numbers strictly less than the parameter $[[i]]$, which
represents the current universe level.  In \rref{I-Set}, function
$[[I]]$ is used to define the meaning of
universes that are strictly smaller than the current level $[[i]]$. The
restriction $[[j < i]]$ in \rref{I-Set} is crucial for our system to
be predicative and we will explain the reason shortly.

@@

To derive consistency, it suffices to restrict $[[S]]$ to the set of
terms that reduce to closed terms. \Rref{I-Void, I-Bool} are
unsurprising; when considering only closed terms, the empty type
should not be inhabited and therefore corresponds to the empty set,
whereas the boolean type is semantically inhabited by terms that
evaluate to the boolean values $[[true]]$ or $[[false]]$. \Rref{I-Eq}
says that an equality type $[[a ~ b : A]]$ corresonds to the
set of terms that evaluate to $[[refl]]$ when $[[a <=> b]]$ holds and
otherwise corresponds to the empty set. Side conditions like $[[a <=>
b]]$ are typically required for indexed types, of which equality types
are an instance.

\Rref{I-Pi} is the most interesting. The precondition consists of a
mysterious function $[[F]]$ that takes an element from the set $[[S]]$, the interpretation
of the type $[[A]]$, and returns a subset of terms. $[[F]]$ is further
subject to the constraint that $[[forall a, (# a in S implies Interp I
i B { a } F a #)]]$. Ignoring the content below the horizontal bar of
the derivation, the statements from the precondition are equivalent to
the conjunction of the following statements:
\begin{itemize}
\item $[[Interp I i A S]]$
\item $\forall [[a]], \text{ if }[[a in S]]\text{, then } \exists
  [[S0]]\text{, } [[Interp I i B { a } S0 ]] $
\end{itemize}
In general, let $R$ be a binary relation over the sets $[[A]]$ and
$[[B]]$, it easy to verify the following equivalence, which is used to
justify the skolemization process~\citep{skolemization}.
\scw{Is this true constructively? If not, is that ok?}
\[\forall a \in A, \exists b \in B \text{ such that } (a,b) \in R
  \iff \exists F \in A \rightarrow B \text, \forall a \in A, (a, F(a))
  \in R\]
The left-hand side is equivalent to the right-hand side, but has no
mentioning of any functions.
In the context of \rref{I-Pi}, through
the equivalence, we can more easily tell what's happening: the
function type $[[Pi A B]]$ has a semantic interpretation if its input
type $[[A]]$ can be semantically
interpreted as some set $[[S]]$, and for all terms $[[a in S]]$, the
type $[[B {a}]]$, obtained by substituing $[[a]]$ into the output type
$[[B]]$, has a semantic interpretation.

If we look at $[[Interp I i Pi A B { b | forall a, (# a in S implies b
  a in F a #) }]]$, the conclusion of \rref{I-Pi}, we see why the
presentation with an explicit $[[F in S -> PowerSet STm]]$ is
useful. We want to state the fact that $[[B {a}]]$ has a semantic
interpretation of some set $[[S0]]$, but we also want to refer to the
set $[[S0]]$ when we talk about the set that $[[Pi A B]]$ corresponds
to. With the function $[[F]]$, we can use $[[F a]]$ to retrive the
witness in the precondition for $[[a in S]]$.
With the alternative representation that is free of the function $[[F]]$, we
might want to reformulate our logical relation in the following way.
\begin{center}
  \drule[]{I-PiAlt}
\end{center}
Unfortunately, \rref{I-PiAlt} not only violates the syntactic strict
positivity constraint required in proof assistants, but
is genuinely non-monotone when written as an endofunction over the
domain of relations.
Intuitively, the failure of monotonicity stems from the fact
that the witness picked in the precondition is not necessarily the
same witness being referred to in the post condition. While it might
be possible to restrict the domain with additional constraints such as
functionality and inversion properties, we opt for our current
skolemized formulation of \rref{I-Pi} so we immediately obtain a
well-defined inductive relation and a usable induction principle. The
slight disadvantage of \rref{I-Pi} is that we need to construct the
function $[[F]]$ each time we apply it. Therefore, in
Lemma~\ref{lemma:piintroalt}, we show the admissibility of
\rref{I-PiAlt} a posteriori and use that as a more convenient
introduction rule for the rest of our proofs.

@@

First, we prove a family of simple properties, which we refer to as
inversion principles for our logical relation. Given $[[Interp I i A
S]]$ where $[[A]]$ is in some head form such as $[[Bool]]$ or $[[Pi A0
B0]]$, the inversion lemma allows us to say something about the set
$[[S]]$. Its proof is simple, but we sketch out the case for
functions to help readers confirm their understanding of \rref{I-Pi}.
\begin{lemma}[Inversion of the logical relation]
  \label{lemma:interpinv}\leavevmode
  \begin{enumerate}
  \item\footnote{\dotv{semtyping.v}{InterpExt\_Void\_inv}} If $[[Interp I i Void S]]$, then $[[S = emptyset]]$.
  \item\footnote{\dotv{semtyping.v}{InterpExt\_Bool\_inv}} If $[[Interp I i Bool S]]$, then $[[S = { a | a =>+ true \/ a =>+ false   }]]$.
  \item\footnote{\dotv{semtyping.v}{InterpExt\_Eq\_inv}} If $[[Interp I i a ~ b : A S]]$, then $[[S = { p | p =>+ refl , a <=> b  }]]$.
  \item\footnote{\dotv{semtyping.v}{InterpExt\_Fun\_inv}} If $[[Interp I i Pi A B S0]]$, then there exists $[[S]],[[F]]$ such that:
    \begin{itemize}
    \item $[[Interp I i A S ]]$
    \item $[[F in S -> PowerSet STm]]$
    \item $[[forall a, (# a in S implies Interp I i B { a } F a #)]]$
    \item $[[S0 = { b | forall a, (# a in S implies b a in F a #) }]]$
    \end{itemize}
  \item\footnote{\dotv{semtyping.v}{InterpExt\_Univ\_inv}} If $[[Interp I i Set j S]]$, then $[[j < i]]$ and $[[S = I j]]$.
  \end{enumerate}
\end{lemma}
\begin{proof}
  As mentioned earlier, we only show the inversion property for the
  function type.
  We start by inducting over the derivation of $[[Interp I i Pi A B S]]$. There
  are only two possible cases we need to consider.
  \begin{description}
  \item[\Rref{I-Pi}:] Immediate.
  \item[\Rref{I-Red}:] We are given that $[[Interp I i Pi A B S0]]$.
    We know that there exists some $[[A0]]$ and
    $[[B0]]$ such that $[[Pi A B => Pi A0 B0]]$ and $[[Interp I i Pi
    A0 B0 S0]]$. From the
    induction hypothesis, there exists $[[S]]$ and $[[F]]$ such that :
    \begin{itemize}
    \item $[[Interp I i A0 S ]]$
    \item $[[F in S -> PowerSet STm]]$
    \item $[[forall a, (# a in S implies Interp I i B0 { a } F a #)]]$
    \item $[[S0 = { b | forall a, (# a in S implies b a in F a #) }]]$
    \end{itemize}
    By inverting the derivation of $[[Pi A B => Pi A0 B0]]$, we derive $[[A => A0]]$ and
    $[[B => B0]]$. By Lemma~\ref{lemma:parsubst}, we have $[[B {a} => B0 {a} ]]$ for all
    $[[a]]$. As a result, by \rref{I-Red}, the same $[[S]]$ and
    $[[F]]$ satisfies the following properties:
    \begin{itemize}
    \item $[[Interp I i A S ]]$
    \item $[[F in S -> PowerSet STm]]$
    \item $[[forall a, (# a in S implies Interp I i B { a } F a #)]]$
    \end{itemize}
    These properties are exactly the ones from the conclusion.
  \end{description}
\end{proof}

\Rref{I-Red} bakes into the logical relation the backward preservation
property. That is, given $[[Interp I i A S]]$, if $[[B =>+ A]]$, then
$[[Interp I i B S]]$ also holds. The following property shows that
preservation holds in the usual forward direction too.
\begin{lemma}[Preservation of the logical relation\footnote{\dotv{semtyping.v}{InterpExt\_preservation}}]
  \label{lemma:interppreservation}
  If $[[Interp I i A S]]$ and $[[A => B]]$, then $[[Interp I i B S]]$.
\end{lemma}
\begin{proof}
  We carry out the proof by induction over the derivation of $[[Interp
  I i A S]]$.
  \begin{description}
  \item[\Rref{I-Void}:] There exists some $[[B]]$ such that $[[Void =>
    B]]$. By inverting the derivation of $[[Void => B]]$, $[[B]]$ must
    be $[[Void]]$ and the result trivially follows.
  \item[\Rref{I-Bool, I-Set}:] Similar to the case for \rref{I-Void}.
  \item[\Rref{I-Eq}:] We know that $[[Interp I i a ~ b : A { p | p =>+
      refl , a <=> b  }]]$ and, by inverting the derivation of
    parallel reduction, $[[a => a0]]$, $[[b => b0]]$, $[[A => A0]]$
    for some $[[a0]]$, $[[b0]]$, and $[[A0]]$. Our goal is to show
    that $[[Interp I i a0 ~ b0 : A {p | p =>+ refl, a <=> b}]]$. By
    \rref{I-Eq}, we already know that $[[Interp I i a0 ~ b0 : A {p | p
      =>+ refl, a0 <=> b0}]]$ and therefore it suffices to show that
    the sets $[[{p | p =>+ refl, a <=> b}]]$ and $[[{p | p =>+ refl,
      a0 <=> b0}]]$ are equal. Equivalently, it suffices to show that
    $[[a <=> b]]$ if and only if $[[a0 <=> b0]]$. By definition, from
    $[[a => a0]]$ and $[[b => b0]]$, we derive $[[a <=> a0]]$ and $[[b
    <=> b0]]$. The result then immediately follows from the fact that
    coherence is an equivalence relation.
  \item[\Rref{I-Pi}:] There exists $[[S]]$ and $[[F]]$
    such that:
    \begin{itemize}
    \item $[[Interp I i Pi A B { b | forall a, (# a in S implies b a in F
        a #) }]]$
    \item $[[Interp I i A S ]]$
    \item $[[F in S -> PowerSet STm]]$
    \item $[[forall a, (# a in S implies Interp I i B { a } F a #)]]$
    \end{itemize}
    There exists some $[[A0]]$ and $[[B0]]$ such that $[[A => A0]]$ and
    $[[B => B0]]$. Our goal is to show that $[[Interp I i Pi A0 B0 { b | forall a, (# a in S implies b a in F
      a #) }]]$. By \rref{I-Pi}, it suffices to show:
    \begin{itemize}
    \item $[[Interp I i A0 S ]]$
    \item $[[F in S -> PowerSet STm]]$
    \item $[[forall a, (# a in S implies Interp I i B0 { a } F a #)]]$
    \end{itemize}
    Since $[[A => A0]]$ and $[[B {a} => B0 {a}]]$ forall $[[a]]$, the
    latter of which follows from Lemma~\ref{lemma:parsubst}, the above
    conditions follow immediately from the induction hypothesis.
  \item[\Rref{I-Red}:] There exists $[[B0]]$ and $[[S]]$ such that $[[A => B0]]$ and
    $[[Interp I i B0 S]]$. Given an arbitrary $[[B]]$ such that $[[A
    => B]]$, our goal is to show that $[[Interp I i B S]]$. By the diamond
    property of parallel
    reduction (Lemma~\ref{lemma:pardiamond}), there exists some term
